Network Structure, Efficiency, and Performance in WikiProjects

### 2018-04-10 preprint/wp-eff-perf.tex
Add preprint.

\def\year{2018}\relax
%File: formatting-instruction.tex
\documentclass[letterpaper]{article} %DO NOT CHANGE THIS
\usepackage{aaai18}  %Required
\usepackage{times}  %Required
\usepackage{helvet}  %Required
\usepackage{courier}  %Required
\usepackage{url}  %Required
\usepackage{graphicx}  %Required
\frenchspacing  %Required
\setlength{\pdfpagewidth}{8.5in}  %Required
\setlength{\pdfpageheight}{11in}  %Required
%PDF Info Is Required:
  \pdfinfo{
/Title (Network Structure, Efficiency, and Performance in WikiProjects)
/Author Edward L. Platt and Daniel M. Romero}
\setcounter{secnumdepth}{0}

% Document-specific packages
\usepackage{epigraph}
\usepackage[normal]{threeparttable}
\usepackage{graphicx}
\usepackage{bbding}

@@

 \begin{document}
% The file aaai.sty is the style file for AAAI Press 
% proceedings, working notes, and technical reports.
%
\title{Network Structure, Efficiency, and Performance in WikiProjects}
%\author{Anonymized \\
%for\\
%blind review
%\And
%Anonymized \\
%for\\
%blind review
\author{Edward L. Platt \\
University of Michigan\\
Ann Arbor, MI\\
elplatt@umich.edu
\And
Daniel M. Romero\\
University of Michigan\\
Ann Arbor, MI\\
drom@umich.edu
}
\maketitle

\begin{abstract}
The internet has enabled collaborations at a scale never before possible,
but the best practices for organizing such large collaborations are still not clear.
Wikipedia is a visible and successful example of such a collaboration which might offer
insight into what makes large-scale, decentralized collaborations successful.
We analyze the relationship between the structural properties of WikiProject coeditor networks
and the performance and efficiency of those projects.
We confirm the existence of an overall performance-efficiency trade-off,
while observing that some projects are higher than others in both performance
and efficiency,
suggesting the existence factors correlating positively with both.
Namely, we find an association between low-degree coeditor networks
and both high performance and high efficiency.
We also confirm results seen in previous numerical and small-scale lab studies:
higher performance with less skewed node distributions,
and higher performance with shorter path lengths.
We use agent-based models to explore possible mechanisms for
degree-dependent performance and efficiency.
We present a novel local-majority learning strategy designed to satisfy properties
of real-world collaborations.
The local-majority strategy as well as a localized conformity-based strategy
both show degree-dependent performance and efficiency,
but in opposite directions,
suggesting that these factors depend on both network structure and learning strategy.
Our results suggest 
possible benefits to decentralized collaborations made of smaller,
more tightly-knit teams,
and that these benefits may be modulated by the particular learning strategies
in use.
\end{abstract}

@@

In addition to our empirical study,
we use agent-based modeling to examine the consequences of specific
assumptions on networked collaboration.
We model individual behavior using a {\em social learning strategy} that
assumes agents 1. can only access a fraction of the model's state,
2. interact with others who share their concerns,
and 3. integrate their preferences into a single state.
Our model is the first we are aware of to incorporate these assumptions,
which are present across many real-world collaborations,
including Wikipedia.

Our main findings are:
\begin{itemize}
\item Despite an overall performance/efficiency trade-off,
WikiProjects with low-degree coeditor networks tend to have both higher performance and higher efficiency;
\item Short paths are associated with higher performance, consistent with a conformity-based learning strategy;
\item Structural inequality, as measured by degree skewness, is associated with lower performance;
\item Our agent-based model
shows that the efficiency and performance of collaborations can
depend on network degree,
and that the direction of that dependence varies with social learning strategy.
\end{itemize}

@@

{\bf Social Learning.} In {\em networked social learning},
agents are represented by nodes on a network
and can interact only with their neighbors.
Social learning tasks can be divided into cases where agents have {\em generated signals}
(independently noisy estimates of a true value)
and those where agents have {\em interpreted signals}
(solutions based on different selections of available data)
\cite{hong_interpreted_2009}.
The behavior of individual agents is described by their
{\em social learning strategy}.
For generated signals,
a naive Bayesian approach converges to the truth
when all agents have the same degree,
while the speed of convergence depends on the {\em spectral gap}
between the two largest eigenvalues of the network's interaction matrix
\cite{degroot_reaching_1974,golub_naive_2010}.
Complex social learning tasks can also be modeled as the problem
of maximizing an objective function with many local maxima,
referred to as a {\em rugged landscape}
\cite{lazer_network_2007,mason_propagation_2008,mason_collaborative_2012,grim_scientific_2013,barkoczi_social_2016}.
Numerical simulations have shown that efficient networks
(those with short paths between nodes)
can result in faster convergence at the cost of a less optimal solution,
due to less time for exploration
\cite{mason_propagation_2008,grim_scientific_2013}.
However, when conformity-based social learning strategies are used,
efficient networks can sometimes find more optimal solutions than
inefficient ones \cite{barkoczi_social_2016}.
Using an agent-based model, Hong and Page \shortcite{hong_groups_2004} found that
diverse groups can outperform groups composed of the best individual
problem-solvers.

{\bf Lab experiments.} Lab-based experiments on networked collaboration
suggest a complex interaction between network topology and other factors.
While groups of networked human subjects perform very well on
difficult graph-coloring tasks, the best performing network architectures
(e.g., fully-connected vs. small-world) vary
from task to task \cite{kearns_experiments_2012}.
The same studies found that while human subjects tend to perform well on
many networks, they perform worst on self-organized
networks, possibly due to higher structural inequality
(degree skewness).
Similarly, some network topologies are able to reach faster decisions in the
presence of more information, while others show the opposite effect
\cite{kearns_experimental_2006}.
Based on lab experiments, Fowler and Christakis \shortcite{fowler_cooperative_2010}
suggest that individual decisions towards altruism are conditional on their
neighbor's behavior and ``contagious'' up to three degrees away.
Later experiments by Suri and Watts \shortcite{suri_cooperation_2011} confirmed the
existence of conditional altruism,
but concluded that altruism
influences only first-degree neighbors.

@@

For Wikipedia specifically,
Robert and Romero \shortcite{robert_crowd_2015} found that
larger group sizes yield higher article ratings
when the groups are diverse and experienced. Kittur and Kraut found that different types of coordination have a complex
effect on the quality of Wikipedia articles \shortcite{kittur_harnessing_2008}.
Both explicit and implicit coordination result in higher quality articles,
with explicit coordination being especially central in the early life of an
article.
Shaw and Hill \shortcite{shaw_laboratories_2014}
found that behavior in online wiki communities is consistent
with the ``iron law of oligarchy,''
which states that
earlier members of a group will, over time, gain disproportionate
decision-making power and act increasingly out of self-interest rather than
the good of the group \cite{michels_political_1999}.
Similarly, Halfaker et al. \shortcite{halfaker_rise_2013} attributed decreasing
participation on Wikipedia to poor retention of new users.
Looking specifically at Wikipedia policies determined by editor consensus,
Keegan and Fiesler \shortcite{keegan_evolution_2017} found a trend
from flexible rule-making towards less flexible maintenance and deliberation.
Using content analysis,
Morgan et al. \shortcite{morgan_project_2013} found WikiProjects to be
more loosely organized than traditional teams.

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[width=3in]{fig-knitting.png}
\caption{
From Wikipedia {\em Knitting} talk page.
Two WikiProjects have assessed the article as B-class quality.
\label{fig:knitting}
}
\end{center}
\end{figure}

@@

For a WikiProject, efficiency quantifies how much participants can raise the
assessed quality of an article for a fixed amount of work.
We measure work by the number of revisions made.
Quality assessments are made through consensus of the project participants themselves.
Different projects can have different standards and practices for assessing article quality,
so the efficiency is not a measure of how quickly some objective measure of quality improves,
but rather of how quickly the project participants can reach consensus on the improvements that
need to be made and make those improvements.
Because our definition relies on assessment transitions,
we must define efficiency variables for
each of the project-level quality assessments: A, B, and C.
For a particular grade $G$,
we desire our definition of efficiency to meet the following conditions:
\begin{itemize}
\item{Strictly increasing in the number of articles reaching grade $G$ (with revision count fixed);}
\item{Strictly decreasing in the number of revisions (with transition count fixed);}
\item{Independent of WikiProject size: not affected by adding an article having the same efficiency.}
\end{itemize}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=2.5in,height=1.67in]{fig-perf-eff.pdf}
\caption{
WikiProject performance is anticorrelated with B-level efficiency,
with Pearson r of -0.12.
Results are similar for other grade levels.
On average, highly efficient WikiProjects are under-performing,
but when looking at specific WikiProjects,
some are higher than others in both performance and efficiency.
\label{fig:eff-perf}
}
\end{figure}

@@

To study the relationship between network structure, efficiency, and performance,
we model the performance and efficiency of WikiProjects using ordinary least-squares linear regression.
Each WikiProject is a single observation.
The models include each project's coeditor network properties as independent variables.
We also include the following project-level variables to control for confounding factors.
\begin{description}
\item[C-efficiency]
(performance only).
Quantifies how quickly a WikiProject improves articles.
Efficiencies for different grades are highly correlated,
so we include only one.
\item[Connected fraction.]
Fraction of coeditor pairs connected by a path.
\item[Talk fraction.] Fraction of total revisions made to talk pages.
\item[Mean similarity.] Mean Jaccard similarity (by article) with other WikiProjects; a measure of topical complexity.
\item[Mean editors/article.] Mean number of editors collaborating on each article in a WikiProject.
\item[Article count.] Total number of articles in the WikiProject.
\item[Editor count.] Total number of editors working on articles within a WikiProject.
\item[Revision count.] Total number of revisions to articles in a WikiProject.
\item[First assessment.] Timestamp of first assessment; a measure of how long a WikiProject has been active.
\item[Mean article age.] Mean age of articles within a WikiProject.
\end{description}

@@

Without additional constraints,
the above model is missing
several key properties of real-world collaborations.
In designing our agent-based model, we paid attention to the
following properties.
\begin{description}
\item[Limited concern.] Agents are concerned only with a subset of the entire
state when making decisions and determining preferences.
(On Wikipedia, editors typically interact with a small subset of the articles.)
\item[Concern-based network.] Agents interact with other agents who share a
common concern over some subset of the state.
(On Wikipedia, editors interact with others who share interests in the same articles.)
\item[Unknown objective.] Agents rank states in order of preference,
but do not have access to the objective function.
(On Wikipedia, there is no ground truth measure of quality.)
\item[Single source of truth.] At any given time, the system is in a single state
and agent preferences are based on local modifications to that state.
(At any point in time, there is only one current version of Wikipedia.)
\end{description}

@@

\bibliographystyle{aaai}
\bibliography{wp-eff-perf}
\end{document}

>>>>

\def\year{2018}\relax
\documentclass[letterpaper,twocolumn,10pt]{article}
\usepackage{times}
\usepackage{helvet}
\usepackage{courier}
\usepackage{url}
\usepackage{graphicx}
\frenchspacing
\setlength{\pdfpagewidth}{8.5in}
\setlength{\pdfpageheight}{11in}
\pdfinfo{
/Title (Network Structure, Efficiency, and Performance in WikiProjects)
/Author (Edward L. Platt, Daniel M. Romero)}
\setcounter{secnumdepth}{0}

% Document-specific packages
\usepackage{epigraph}
\usepackage[normal]{threeparttable}
\usepackage{graphicx}
\usepackage{bbding}
\usepackage{natbib}
\usepackage[letterpaper,left=0.75in,right=0.75in,top=0.75in,bottom=1.25in]{geometry}
\usepackage{titlesec}

@@

\setlength{\columnsep}{0.375in}
\titleformat*{\section}{\large\bfseries}
\titleformat*{\subsection}{\normalsize\bfseries}
\titleformat*{\subsubsection}{\normalsize\bfseries}
\titleformat*{\paragraph}{\normalsize\bfseries}
\titleformat*{\subparagraph}{\normalsize\bfseries}
\titlespacing\section{0pt}{12pt plus 4pt minus 2pt}{0pt plus 2pt minus 2pt}
\titlespacing\subsection{0pt}{12pt plus 4pt minus 2pt}{0pt plus 2pt minus 2pt}

\linespread{1.0}

\begin{document}
\title{Network Structure, Efficiency, and Performance in WikiProjects}
\author{
Edward L. Platt \\
University of Michigan\\
elplatt@umich.edu
\and
Daniel M. Romero\\
University of Michigan\\
drom@umich.edu
}
\date{}
\maketitle

\begin{abstract}
\small
The internet has enabled collaborations at a scale never before possible,
but the best practices for organizing such large collaborations are still not clear.
Wikipedia is a visible and successful example of such a collaboration which might offer
insight into what makes large-scale, decentralized collaborations successful.
We analyze the relationship between the structural properties of WikiProject coeditor networks
and the performance and efficiency of those projects.
We confirm the existence of an overall performance-efficiency trade-off,
while observing that some projects are higher than others in both performance
and efficiency,
suggesting the existence factors correlating positively with both.
Namely, we find an association between low-degree coeditor networks
and both high performance and high efficiency.
We also confirm results seen in previous numerical and small-scale lab studies:
higher performance with less skewed node distributions,
and higher performance with shorter path lengths.
We use agent-based models to explore possible mechanisms for
degree-dependent performance and efficiency.
We present a novel local-majority learning strategy designed to satisfy properties
of real-world collaborations.
The local-majority strategy as well as a localized conformity-based strategy
both show degree-dependent performance and efficiency,
but in opposite directions,
suggesting that these factors depend on both network structure and learning strategy.
Our results suggest 
possible benefits to decentralized collaborations made of smaller,
more tightly-knit teams,
and that these benefits may be modulated by the particular learning strategies
in use.
\end{abstract}

@@

In addition to our empirical study,
we use agent-based modeling to examine the consequences of specific
assumptions on networked collaboration.
We model individual behavior using a {\em social learning strategy} that
assumes agents 1. can only access a fraction of the model's state,
2. interact with others who share their concerns,
and 3. integrate their preferences into a single state.
Our model is the first we are aware of to incorporate these assumptions,
which are present across many real-world collaborations,
including Wikipedia.
\pagebreak

Our main findings are:
\begin{itemize}
\setlength\itemsep{0pt}
\item Despite an overall performance/efficiency trade-off,
WikiProjects with low-degree coeditor networks tend to have both higher performance and higher efficiency;
\item Short paths are associated with higher performance, consistent with a conformity-based learning strategy;
\item Structural inequality, as measured by degree skewness, is associated with lower performance;
\item Our agent-based model
shows that the efficiency and performance of collaborations can
depend on network degree,
and that the direction of that dependence varies with social learning strategy.
\end{itemize}

@@

{\bf Social Learning.} In {\em networked social learning},
agents are represented by nodes on a network
and can interact only with their neighbors.
Social learning tasks can be divided into cases where agents have {\em generated signals}
(independently noisy estimates of a true value)
and those where agents have {\em interpreted signals}
(solutions based on different selections of available data)
\cite{hong_interpreted_2009}.
The behavior of individual agents is described by their
{\em social learning strategy}.
For generated signals,
a naive Bayesian approach converges to the truth
when all agents have the same degree,
while the speed of convergence depends on the {\em spectral gap}
between the two largest eigenvalues of the network's interaction matrix
\cite{degroot_reaching_1974,golub_naive_2010}.
Complex social learning tasks can also be modeled as the problem
of maximizing an objective function with many local maxima,
referred to as a {\em rugged landscape}
\cite{lazer_network_2007,mason_propagation_2008,mason_collaborative_2012,grim_scientific_2013,barkoczi_social_2016}.
Numerical simulations have shown that efficient networks
(those with short paths between nodes)
can result in faster convergence at the cost of a less optimal solution,
due to less time for exploration
\cite{mason_propagation_2008,grim_scientific_2013}.
However, when conformity-based social learning strategies are used,
efficient networks can sometimes find more optimal solutions than
inefficient ones \cite{barkoczi_social_2016}.
Using an agent-based model, Hong and Page \citeyear{hong_groups_2004} found that
diverse groups can outperform groups composed of the best individual
problem-solvers.

{\bf Lab experiments.} Lab-based experiments on networked collaboration
suggest a complex interaction between network topology and other factors.
While groups of networked human subjects perform very well on
difficult graph-coloring tasks, the best performing network architectures
(e.g., fully-connected vs. small-world) vary
from task to task \cite{kearns_experiments_2012}.
The same studies found that while human subjects tend to perform well on
many networks, they perform worst on self-organized
networks, possibly due to higher structural inequality
(degree skewness).
Similarly, some network topologies are able to reach faster decisions in the
presence of more information, while others show the opposite effect
\cite{kearns_experimental_2006}.
Based on lab experiments, Fowler and Christakis \citeyear{fowler_cooperative_2010}
suggest that individual decisions towards altruism are conditional on their
neighbor's behavior and ``contagious'' up to three degrees away.
Later experiments by Suri and Watts \citeyear{suri_cooperation_2011} confirmed the
existence of conditional altruism,
but concluded that altruism
influences only first-degree neighbors.

@@

For Wikipedia specifically,
Robert and Romero \citeyear{robert_crowd_2015} found that
larger group sizes yield higher article ratings
when the groups are diverse and experienced. Kittur and Kraut found that different types of coordination have a complex
effect on the quality of Wikipedia articles \citeyear{kittur_harnessing_2008}.
Both explicit and implicit coordination result in higher quality articles,
with explicit coordination being especially central in the early life of an
article.
Shaw and Hill \citeyear{shaw_laboratories_2014}
found that behavior in online wiki communities is consistent
with the ``iron law of oligarchy,''
which states that
earlier members of a group will, over time, gain disproportionate
decision-making power and act increasingly out of self-interest rather than
the good of the group \cite{michels_political_1999}.
Similarly, Halfaker et al. \citeyear{halfaker_rise_2013} attributed decreasing
participation on Wikipedia to poor retention of new users.
Looking specifically at Wikipedia policies determined by editor consensus,
Keegan and Fiesler \citeyear{keegan_evolution_2017} found a trend
from flexible rule-making towards less flexible maintenance and deliberation.
Using content analysis,
Morgan et al. \citeyear{morgan_project_2013} found WikiProjects to be
more loosely organized than traditional teams.

@@

\begin{figure}[t!]
\begin{center}
\includegraphics[width=3in]{fig-knitting.png}
\caption{
From Wikipedia {\em Knitting} talk page.
Two WikiProjects have assessed the article as B-class quality.
\label{fig:knitting}
}
\end{center}
\end{figure}

@@

For a WikiProject, efficiency quantifies how much participants can raise the
assessed quality of an article for a fixed amount of work.
We measure work by the number of revisions made.
Quality assessments are made through consensus of the project participants themselves.
Different projects can have different standards and practices for assessing article quality,
so the efficiency is not a measure of how quickly some objective measure of quality improves,
but rather of how quickly the project participants can reach consensus on the improvements that
need to be made and make those improvements.
Because our definition relies on assessment transitions,
we must define efficiency variables for
each of the project-level quality assessments: A, B, and C.
For a particular grade $G$,
we desire our definition of efficiency to meet the following conditions:
\begin{itemize}
\setlength\itemsep{0pt}
\item{Strictly increasing in the number of articles reaching grade $G$ (with revision count fixed);}
\item{Strictly decreasing in the number of revisions (with transition count fixed);}
\item{Independent of WikiProject size: not affected by adding an article having the same efficiency.}
\end{itemize}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=2.5in,height=1.67in]{fig-perf-eff.pdf}
\caption{
WikiProject performance is anticorrelated with B-level efficiency,
with Pearson r of -0.12.
Results are similar for other grade levels.
On average, highly efficient WikiProjects are under-performing,
but when looking at specific WikiProjects,
some are higher than others in both performance and efficiency.
\label{fig:eff-perf}
}
\end{figure}

@@

To study the relationship between network structure, efficiency, and performance,
we model the performance and efficiency of WikiProjects using ordinary least-squares linear regression.
Each WikiProject is a single observation.
The models include each project's coeditor network properties as independent variables.
We also include the following project-level variables to control for confounding factors.
\begin{description}
\setlength\itemsep{0pt}
\item[C-efficiency]
(performance only).
Quantifies how quickly a WikiProject improves articles.
Efficiencies for different grades are highly correlated,
so we include only one.
\item[Connected fraction.]
Fraction of coeditor pairs connected by a path.
\item[Talk fraction.] Fraction of total revisions made to talk pages.
\item[Mean similarity.] Mean Jaccard similarity (by article) with other WikiProjects; a measure of topical complexity.
\item[Mean editors/article.] Mean number of editors collaborating on each article in a WikiProject.
\item[Article count.] Total number of articles in the WikiProject.
\item[Editor count.] Total number of editors working on articles within a WikiProject.
\item[Revision count.] Total number of revisions to articles in a WikiProject.
\item[First assessment.] Timestamp of first assessment; a measure of how long a WikiProject has been active.
\item[Mean article age.] Mean age of articles within a WikiProject.
\end{description}

@@

Without additional constraints,
the above model is missing
several key properties of real-world collaborations.
In designing our agent-based model, we paid attention to the
following properties.
\begin{description}
\setlength\itemsep{0pt}
\item[Limited concern.] Agents are concerned only with a subset of the entire
state when making decisions and determining preferences.
(On Wikipedia, editors typically interact with a small subset of the articles.)
\item[Concern-based network.] Agents interact with other agents who share a
common concern over some subset of the state.
(On Wikipedia, editors interact with others who share interests in the same articles.)
\item[Unknown objective.] Agents rank states in order of preference,
but do not have access to the objective function.
(On Wikipedia, there is no ground truth measure of quality.)
\item[Single source of truth.] At any given time, the system is in a single state
and agent preferences are based on local modifications to that state.
(At any point in time, there is only one current version of Wikipedia.)
\end{description}

@@

\bibliographystyle{apalike}
\bibliography{wp-eff-perf}

\end{document}
